A Beginners Guide to Perfomance Factors of Machine Learning Applications with Focus on Neural Nets and Deep Learning

### 2018-01-09 paper.tex
some changes conclusion and challenges

\author{\IEEEauthorblockN{Johannes Wünsche}
\IEEEauthorblockA{Otto-von-Guericke University Magdeburg\\
Email: johannes.wuensche@st.ovgu.de}
}

@@

\IEEEpeerreviewmaketitle
\section{Introduction}
Machine Learning seems to be everywhere around us. With the recent success of Deep, Convolutional,and Neural Networks, and other Machine Learning tools, especially the, in general media, well reported Learners like AlphaGo, Silver or Watson, these are experiencing a kind of renaissance their usage has spread from industrial(e.g. Self-driving cars rely on autonomous interpretation of camera feed and sensor output that can be optimized by using machine learning) to retail. But also in modern automation and research, machine learning becomes a greater field managing to find more efficient ways to achieve better results in many expertise.

@@

As a small scale neural network applications we define a network with few neural nodes inside the network and because of this rather limited size most of the time run on a single machine and or single processing unit(worker) which can complete the training set in a reasonable time.

@@

On the other hand as a large scale neural network applications we define a network with a large number of nodes which can because of the complexity of the resulting connections take a large amount of time to complete training if computed on one worker and therefore will take use of a distributed system consisting of multiple workers and one or more servers to synchronize the workers and distribute the required data.

@@

\section{Choice of Processing Units}
The next important choice is the choice of the actual operating calculation unit. We distinguish them into three main models, CPU, GPU and GPU-Clustering.
\subsection{CPU}
As the first most naive option we have a look on the CPU for training of the neural net with the simple single thread and more complex multi-thread calculation.
\subsubsection{Single-thread}
The most naive implementation is the single-thread CPU calculation, because oft this more  brute force approach and sequential execution and calculation this is also the slowest and with a larger number of connected nodes simply unfeasible because of the excessive calculation time required . But as can be thought of this does not uitilize the full capacity of the CPU and therefore results in performance loss in comparison to multithread usage(Table I).
\subsubsection{Multi-thread}
A more complex approach consist of utilizing the multi-core characterisitics of modern CPUs, by splitting of training calculation with only requirements to already calculated values. This is for example preferrable when training with the help of a \emph{Backpropagation} algorithm. The effect  continues to grow by using more threads but reaches it limits when using more threads than physical cores available.
\begin{table} 
\centering
\begin{tabular}{c c c c c}
\hline
Framework & 1 Thread & 2 Threads &4 Threads & 8 Threads\\\hline
Caffe & 1.324 & 0.790 & 0.578 & 15.444 \\
Tensorflow & 7.062 & 4.789 & 2.648 & 1.938 \\
Torch & 1.329 & 0.710 & 0.423 & na \\\hline
\end{tabular}
\caption{Training time(in s) for a mini-batch of size 64 of a Fully Connected Network(FCN) trained with synthetic data on a i7-3820 with 4-physical cores using 3 different commonly used Deep Learning frameworks\cite{shi2016benchmarking}}
\label{fig_ttfcn}
\end{table}
Though this approach is limited by the actual core number available and should not overstep the number of physical cores, like with intel \emph{hyper-threading} technology, which can lengthen the required calculation time because of a more inefficient usage\cite{shi2016benchmarking}.
\subsection{GPU}
A more advanced implementation is to use the shared memory, multi-core environment of modern GPUs. For example with NVIDIAs CUDA this can be done efficiently and result in a speed-up of calculation of up to 3 times the CPU-based approachs performance for deep learning algorithms over optimized floting-point baseline\cite{shi2016benchmarking}. Almost all recent neural net learning frameworks support this calculation unit because of its excellent performance rating for neural networks.

@@

To ease the effect of GPU to GPU communication,by relieveing the CPU of managing the , technologies like GPUDirect from NVIDIA\cite{nvidiagpudirect2017} can be used allowing direct GPU to GPU communication if they are connected to the same PCIe bus, so this is only appplicable within a GPU Cluster or a single system and does not remove all communication overhead

@@

\section{Conclusion}
%Summary
During our survey we observed that the most influencing factors are the basic choice of neural network to be used, this includes its complexity as well as the actual architecture the network is build of. Further the processing unit that is chosen has a strong influence on the quality and speed the final network that lies within acceptable error limits will be reached. Based on the choice of this other factors come into consideration influencing the process like communication between processing units and parallelism of data between them. They are not quite as influential as the first two but can lead to a strong improvement or worsening.

%Takeaway
Especially data transfer time and alike carries a potential to lengthen the training time of an significant amount. To ease this technology like GPUDirect\cite{nvidiagpudirect2017} and data prefetching\cite{yang2010gpgpu} can be used. It may be possible to decrease the effect of the data flow further by introducing a direct memory transfer unit which can indepently organize data and can handle parallel transfers to and from multiple units. This could decrease the calculation time to a bare minimum but would requires additional hardware and maybe modification of the used GPU and CPU.  Although this would break security of the system on broder range because of the unauthorized memory access and changes, it could offer a steep improvement for large calculations on distributed systems.

TODO: Takeaway, different ideas, innnovative

>>>>

\author{\IEEEauthorblockN{Johannes Wünsche}
\IEEEauthorblockA{Otto-von-Guericke University Magdeburg\\
johannes.wuensche@st.ovgu.de}
}

@@

\IEEEpeerreviewmaketitle
\section{Introduction}
Machine Learning seems to be everywhere around us. With the recent success of Neural Networks, and other Machine Learning tools, especially the, in general media, well reported learners like AlphaGo, Silver or IBM Watson, these are experiencing a kind of renaissance their usage has spread from industrial(e.g. Self-driving cars rely on autonomous interpretation of camera feed and sensor output that can be optimized by using machine learning) to retail. But also in modern automation and research, machine learning becomes a greater field managing to find more efficient ways to achieve better results in many expertise.

@@

As a small scale neural network applications we define a network with few neural nodes inside the network and because of this rather limited size it is most of the time run on a single machine and or single processing unit which can complete the training process of the net in a reasonable time.

@@

On the other hand as a large scale neural network applications we define a network with a large number of nodes which can because of the complexity of the resulting connections take a large amount of time to complete training if computed on one processing unit and therefore will take use of a distributed system consisting of multiple units in a single system or more servers to synchronize multiple systems of processing units and distribute the required data.

@@

\section{Choice of Processing Units}
The next important choice is the choice of the actual operating calculation unit. We distinguish them into three main models, CPU, GPU and GPU-Clustering.
\subsection{CPU}
As the first most naive option we have a look on the CPU for training of the neural net with the simple single thread and more complex multi-thread calculation.
\subsubsection{Single-thread}
The most naive implementation is the single-thread CPU calculation, because oft this more  brute force approach and sequential execution and calculation this is also the slowest and with a larger number of connected nodes simply unfeasible because of the excessive calculation time required . But as can be thought of this does not uitilize the full capacity of the CPU and therefore results in performance loss in comparison to multithread usage(Table I).
\subsubsection{Multi-thread}
A more complex approach consist of utilizing the multi-core characterisitics of modern CPUs, by splitting of training calculation with only requirements to already calculated values. This is for example preferrable when training with the help of a \emph{Backpropagation} algorithm. The effect  continues to grow by using more threads but reaches it limits when using more threads than physical cores available.
\begin{table} 
\centering
\begin{tabular}{c c c c c}
\hline
Framework & 1 Thread & 2 Threads &4 Threads & 8 Threads\\\hline
Caffe & 1.324 & 0.790 & 0.578 & 15.444 \\
Tensorflow & 7.062 & 4.789 & 2.648 & 1.938 \\
Torch & 1.329 & 0.710 & 0.423 & na \\\hline
\end{tabular}
\caption{Training time(in s) for a mini-batch of size 64 of a Fully Connected Network(FCN) trained with synthetic data on a i7-3820 with 4-physical cores using 3 different commonly used Deep Learning frameworks\cite{shi2016benchmarking}}
\label{fig_ttfcn}
\end{table}
Though this approach is limited by the actual core number available and should not overstep the number of physical cores, like with intel \emph{hyper-threading} technology, which can lengthen the required calculation time because of a more inefficient usage\cite{shi2016benchmarking}.
\subsection{GPU}
A more advanced implementation is to use the shared memory, multi-core environment of modern GPUs. For example with the help of  NVIDIAs CUDA this can be done efficiently and result in a speed-up of calculation of up to 3 times the CPU-based approachs performance for deep learning algorithms over optimized floting-point baseline\cite{shi2016benchmarking}. Almost all recent neural net learning frameworks support GPUs because of its excellent performance rating for neural networks.

@@

To ease the effect of GPU to GPU communication,by relieveing the CPU of managing them, technologies like GPUDirect from NVIDIA\cite{nvidiagpudirect2017} can be used allowing direct GPU to GPU communication if they are connected to the same PCIe bus, so this is only appplicable within a GPU Cluster or a single system and does not remove all communication overhead. Also shared memory between GPUs can come in place saving synchronization between them.

@@

\section{Conclusion}
%Summary
During our survey we observed that the most influencing factors are the basic choice of neural network to be used, this includes its complexity as well as the actual architecture the network is build of. Further the processing unit that is chosen has a strong influence on the quality and speed the final network that lies within acceptable error limits will be reached. Based on the choice of this other factors came into consideration influencing the process, like communication between processing units and parallelism of data between them. They are not quite as influential as the first two but can lead to a strong improvement or worsening.

%Takeaway
Especially data transfer time and alike carries a potential to lengthen the training time of an significant amount. To ease this technology like GPUDirect\cite{nvidiagpudirect2017} and data prefetching\cite{yang2010gpgpu} can be used. It may be possible to decrease the effect of the data flow further by introducing a direct memory transfer unit which can indepently organize synchronization and can handle parallel transfers to and from multiple GPUs. This could decrease the calculation time to a bare minimum but would require additional hardware and maybe modification of the used GPUs. Although this could break some security limits because of unauthorized memory accesses.
